---
title: "blendScenes Deep Features Report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(readxl)
library(tidyr)

workbook_path <- "../data/formr_downloads/blendScenes_prolific_A/results_version_A_long_with_images_and_paths.xlsx"
feature_dir <- "../data/deep_features/version_A_places365_alexnet"
blends_path <- "../data/Blends.xlsx"

normalize_image_id <- function(x) {
  x <- as.character(x)
  x <- sub("\\.[A-Za-z0-9]+$", "", x)
  trimws(x)
}

results_long <- read_excel(workbook_path, sheet = "long_results")
blends_raw <- read_excel(blends_path)

results_long <- results_long |>
  filter(trial_type == "trial") |>
  mutate(
    rating = as.numeric(rating),
    blend_image_id = normalize_image_id(blend_image),
    left_image_id = normalize_image_id(left_image),
    right_image_id = normalize_image_id(right_image),
    behavior_right = (rating - 1) / 99,
    behavior_left = 1 - behavior_right
  )

behavior_trial <- results_long |>
  group_by(trial_id, blend_image, left_image, right_image, blend_image_id, left_image_id, right_image_id) |>
  summarise(
    n_ratings = sum(!is.na(rating)),
    mean_rating = mean(rating, na.rm = TRUE),
    mean_right = mean(behavior_right, na.rm = TRUE),
    mean_left = mean(behavior_left, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(behavior_diff_right_minus_left = mean_right - mean_left)

blend_lookup <- blends_raw |>
  transmute(
    trial_id = as.integer(id_blendu),
    category1 = as.character(category1),
    category2 = as.character(category2),
    category_pair = paste0(category1, "-", category2)
  )

cosine_pair <- function(a, b) {
  denom <- sqrt(sum(a * a)) * sqrt(sum(b * b))
  if (is.na(denom) || denom == 0) {
    return(NA_real_)
  }
  sum(a * b) / denom
}

l2_pair <- function(a, b) {
  sqrt(sum((a - b) ^ 2))
}

feature_files <- c("conv3_features.csv", "conv5_features.csv", "fc6_features.csv", "fc7_features.csv")

feature_trial_list <- lapply(feature_files, function(file_name) {
  layer_name <- sub("_features\\.csv$", "", basename(file_name))
  feature_df <- read_csv(file.path(feature_dir, file_name), show_col_types = FALSE)
  feature_df <- feature_df |>
    mutate(image_id = normalize_image_id(image_name))

  feature_mat <- feature_df |>
    select(-image_name, -image_id) |>
    as.matrix()
  rownames(feature_mat) <- feature_df$image_id

  trial_sim <- behavior_trial |>
    rowwise() |>
    mutate(
      layer = layer_name,
      sim_blend_left = cosine_pair(feature_mat[blend_image_id, ], feature_mat[left_image_id, ]),
      sim_blend_right = cosine_pair(feature_mat[blend_image_id, ], feature_mat[right_image_id, ]),
      l2_blend_left = l2_pair(feature_mat[blend_image_id, ], feature_mat[left_image_id, ]),
      l2_blend_right = l2_pair(feature_mat[blend_image_id, ], feature_mat[right_image_id, ])
    ) |>
    ungroup() |>
    mutate(
      sim_diff_right_minus_left = sim_blend_right - sim_blend_left,
      sim_sum = sim_blend_right + sim_blend_left,
      l2_diff_right_minus_left = l2_blend_right - l2_blend_left
    )

  trial_sim
})

feature_trial <- bind_rows(feature_trial_list) |>
  left_join(blend_lookup, by = "trial_id")

layer_summary <- feature_trial |>
  group_by(layer) |>
  summarise(
    n_trials = n(),
    cor_cosine_diff_behavior = cor(sim_diff_right_minus_left, behavior_diff_right_minus_left, use = "complete.obs"),
    cor_l2_diff_behavior = cor(l2_diff_right_minus_left, behavior_diff_right_minus_left, use = "complete.obs"),
    cor_right_cosine_mean_right = cor(sim_blend_right, mean_right, use = "complete.obs"),
    cor_left_cosine_mean_left = cor(sim_blend_left, mean_left, use = "complete.obs"),
    cor_right_l2_mean_right = cor(l2_blend_right, mean_right, use = "complete.obs"),
    cor_left_l2_mean_left = cor(l2_blend_left, mean_left, use = "complete.obs"),
    cosine_lm_r2 = summary(lm(behavior_diff_right_minus_left ~ sim_diff_right_minus_left))$r.squared,
    l2_lm_r2 = summary(lm(behavior_diff_right_minus_left ~ l2_diff_right_minus_left))$r.squared,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

overall_summary <- tibble(
  metric = c(
    "Observed version A trials",
    "Participants per trial",
    "Feature layers summarized",
    "Behavior measure",
    "Cosine measure",
    "L2 measure"
  ),
  value = c(
    n_distinct(feature_trial$trial_id),
    paste(range(behavior_trial$n_ratings), collapse = "-"),
    paste(sort(unique(feature_trial$layer)), collapse = ", "),
    "Mean slider mapped to right preference: (rating - 1) / 99",
    "Cosine similarity between blend and left/right image feature vectors",
    "Euclidean distance between blend and left/right image feature vectors"
  )
)
```

## Overview

This report links behavioral judgments from [results_version_A_long_with_images_and_paths.xlsx](`r workbook_path`) to deep-feature similarity from the Places365 AlexNet feature exports in [version_A_places365_alexnet](`r feature_dir`).

For each trial, the report computes:

- `sim_blend_left`: cosine similarity between the blend image and the left image
- `sim_blend_right`: cosine similarity between the blend image and the right image
- `sim_diff_right_minus_left`: right similarity minus left similarity
- `l2_blend_left`: Euclidean distance between the blend image and the left image
- `l2_blend_right`: Euclidean distance between the blend image and the right image
- `l2_diff_right_minus_left`: right distance minus left distance
- `behavior_diff_right_minus_left`: average behavioral pull to the right minus left, derived from the slider

```{r}
kable(overall_summary, col.names = c("Metric", "Value"))
```

## Layer Summary

```{r}
kable(layer_summary, col.names = c("Layer", "Trials", "Cor: cosine diff vs behavior diff", "Cor: L2 diff vs behavior diff", "Cor: right cosine vs mean right", "Cor: left cosine vs mean left", "Cor: right L2 vs mean right", "Cor: left L2 vs mean left", "Cosine model R2", "L2 model R2"))
```

## Trial-Level Table

```{r}
trial_table <- feature_trial |>
  select(layer, trial_id, category_pair, blend_image, left_image, right_image, n_ratings, mean_rating, mean_left, mean_right, sim_blend_left, sim_blend_right, sim_diff_right_minus_left, l2_blend_left, l2_blend_right, l2_diff_right_minus_left, behavior_diff_right_minus_left) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(trial_table)
```

## Cosine Similarity vs Behavior

```{r, fig.width=8, fig.height=4.8}
ggplot(feature_trial, aes(x = sim_diff_right_minus_left, y = behavior_diff_right_minus_left, color = layer)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    x = "Feature similarity difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "Trial-level relation between feature similarity and behavioral choice"
  ) +
  theme_minimal(base_size = 11)
```

## Cosine By Layer

```{r, fig.width=10, fig.height=4.8}
ggplot(feature_trial, aes(x = sim_diff_right_minus_left, y = behavior_diff_right_minus_left)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.8, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, color = "#D95F0E") +
  facet_wrap(~ layer, scales = "free_x") +
  labs(
    x = "Feature similarity difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "Relation split by feature layer"
  ) +
  theme_minimal(base_size = 11)
```

## Cosine Left and Right Separately

```{r, fig.width=10, fig.height=4.8}
plot_df <- feature_trial |>
  select(layer, trial_id, mean_left, mean_right, sim_blend_left, sim_blend_right) |>
  pivot_longer(
    cols = c(mean_left, mean_right, sim_blend_left, sim_blend_right),
    names_to = c(".value", "side"),
    names_pattern = "(mean|sim_blend)_(left|right)"
  )

ggplot(plot_df, aes(x = sim_blend, y = mean, color = side)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ layer) +
  labs(
    x = "Blend-to-side cosine similarity",
    y = "Average behavioral preference for that side",
    title = "Side-specific similarity and side-specific behavioral preference"
  ) +
  theme_minimal(base_size = 11)
```

## L2 Distance vs Behavior

```{r, fig.width=8, fig.height=4.8}
ggplot(feature_trial, aes(x = l2_diff_right_minus_left, y = behavior_diff_right_minus_left, color = layer)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    x = "Feature L2 distance difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "Trial-level relation between L2 distance and behavioral choice"
  ) +
  theme_minimal(base_size = 11)
```

## L2 By Layer

```{r, fig.width=10, fig.height=4.8}
ggplot(feature_trial, aes(x = l2_diff_right_minus_left, y = behavior_diff_right_minus_left)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.8, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, color = "#D95F0E") +
  facet_wrap(~ layer, scales = "free_x") +
  labs(
    x = "Feature L2 distance difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "Relation split by feature layer using L2 distance"
  ) +
  theme_minimal(base_size = 11)
```

## L2 Left and Right Separately

```{r, fig.width=10, fig.height=4.8}
l2_plot_df <- feature_trial |>
  select(layer, trial_id, mean_left, mean_right, l2_blend_left, l2_blend_right) |>
  pivot_longer(
    cols = c(mean_left, mean_right, l2_blend_left, l2_blend_right),
    names_to = c(".value", "side"),
    names_pattern = "(mean|l2_blend)_(left|right)"
  )

ggplot(l2_plot_df, aes(x = l2_blend, y = mean, color = side)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ layer, scales = "free_x") +
  labs(
    x = "Blend-to-side Euclidean distance",
    y = "Average behavioral preference for that side",
    title = "Side-specific L2 distance and side-specific behavioral preference"
  ) +
  theme_minimal(base_size = 11)
```

## By Blend Type

```{r}
pair_summary <- feature_trial |>
  group_by(layer, category_pair) |>
  summarise(
    n_trials = n(),
    cor_cosine_diff_behavior = cor(sim_diff_right_minus_left, behavior_diff_right_minus_left, use = "complete.obs"),
    cor_l2_diff_behavior = cor(l2_diff_right_minus_left, behavior_diff_right_minus_left, use = "complete.obs"),
    cosine_lm_r2 = summary(lm(behavior_diff_right_minus_left ~ sim_diff_right_minus_left))$r.squared,
    l2_lm_r2 = summary(lm(behavior_diff_right_minus_left ~ l2_diff_right_minus_left))$r.squared,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(pair_summary, col.names = c("Layer", "Blend Type", "Trials", "Cor: cosine diff vs behavior diff", "Cor: L2 diff vs behavior diff", "Cosine model R2", "L2 model R2"))
```

## Cosine by Blend Type

```{r, fig.width=12, fig.height=7}
ggplot(feature_trial, aes(x = sim_diff_right_minus_left, y = behavior_diff_right_minus_left, color = category_pair)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(category_pair ~ layer, scales = "free") +
  labs(
    x = "Feature similarity difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "Cosine similarity and behavior split by blend type and layer"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

## L2 by Blend Type

```{r, fig.width=12, fig.height=7}
ggplot(feature_trial, aes(x = l2_diff_right_minus_left, y = behavior_diff_right_minus_left, color = category_pair)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(category_pair ~ layer, scales = "free") +
  labs(
    x = "Feature L2 distance difference: right minus left",
    y = "Behavioral preference difference: right minus left",
    title = "L2 distance and behavior split by blend type and layer"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```
