---
title: "blendScenes Results Report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(jsonlite)
library(ggplot2)
library(knitr)
library(readxl)

results_path <- "../data/formr_downloads/blendScenes_prolific_A/results.json"
blends_path <- "../data/Blends.xlsx"
images_dir <- "report_images"

results_raw <- fromJSON(results_path)
blends_raw <- read_excel(blends_path)

trial_cols <- grep("^slider_[0-9]+$", names(results_raw), value = TRUE)
catch_cols <- grep("^slider_catch[0-9]+$", names(results_raw), value = TRUE)

trial_ids <- as.integer(sub("^slider_", "", trial_cols))
trial_matrix <- as.data.frame(results_raw[trial_cols], check.names = FALSE)
trial_matrix[] <- lapply(trial_matrix, as.numeric)

catch_matrix <- as.data.frame(results_raw[catch_cols], check.names = FALSE)
catch_matrix[] <- lapply(catch_matrix, as.numeric)

long_trials <- do.call(
  rbind,
  lapply(seq_along(trial_cols), function(i) {
    data.frame(
      session_id = results_raw$session_id,
      iteration = results_raw$iteration,
      prolific_code = results_raw$prolific_code,
      trial_id = trial_ids[i],
      rating = trial_matrix[[i]]
    )
  })
)

long_trials <- long_trials[!is.na(long_trials$rating), ]
long_trials <- long_trials[order(long_trials$trial_id, long_trials$session_id), ]
rownames(long_trials) <- NULL

observed_trial_summary <- do.call(
  rbind,
  lapply(split(long_trials$rating, long_trials$trial_id), function(x) {
    data.frame(
      n = length(x),
      mean = mean(x),
      sd = sd(x),
      median = median(x),
      min = min(x),
      max = max(x)
    )
  })
)

observed_trial_summary$trial_id <- as.integer(rownames(observed_trial_summary))
rownames(observed_trial_summary) <- NULL

blend_lookup <- data.frame(
  trial_id = as.integer(blends_raw$id_blendu),
  category1 = as.character(blends_raw$category1),
  category2 = as.character(blends_raw$category2),
  category_pair = paste0(blends_raw$category1, "-", blends_raw$category2),
  img_name1 = as.character(blends_raw$img_name1),
  img_name2 = as.character(blends_raw$img_name2),
  blend_name = as.character(blends_raw$blend_name)
)

observed_trial_summary$image_file <- file.path(images_dir, paste0(observed_trial_summary$trial_id, "_kolaz.png"))
observed_trial_summary <- merge(observed_trial_summary, blend_lookup, by = "trial_id", all.x = TRUE, sort = TRUE)
observed_trial_summary <- observed_trial_summary[, c("trial_id", "blend_name", "category1", "category2", "category_pair", "img_name1", "img_name2", "n", "mean", "sd", "median", "min", "max", "image_file")]

icc_2way_random_absolute <- function(mat) {
  mat <- as.matrix(mat)
  storage.mode(mat) <- "double"
  if (anyNA(mat)) {
    keep_cols <- colSums(is.na(mat)) == 0
    mat <- mat[, keep_cols, drop = FALSE]
  }
  if (anyNA(mat)) {
    keep_rows <- rowSums(is.na(mat)) == 0
    mat <- mat[keep_rows, , drop = FALSE]
  }
  n <- nrow(mat)
  k <- ncol(mat)
  if (n < 2 || k < 2) {
    return(list(icc21 = NA_real_, icc2k = NA_real_, n_targets = n, n_raters = k))
  }

  grand_mean <- mean(mat)
  row_means <- rowMeans(mat)
  col_means <- colMeans(mat)

  ss_rows <- k * sum((row_means - grand_mean) ^ 2)
  ss_cols <- n * sum((col_means - grand_mean) ^ 2)
  ss_total <- sum((mat - grand_mean) ^ 2)
  ss_error <- ss_total - ss_rows - ss_cols

  ms_rows <- ss_rows / (n - 1)
  ms_cols <- ss_cols / (k - 1)
  ms_error <- ss_error / ((n - 1) * (k - 1))

  icc21 <- (ms_rows - ms_error) / (ms_rows + (k - 1) * ms_error + (k * (ms_cols - ms_error) / n))
  icc2k <- (ms_rows - ms_error) / (ms_rows + ((ms_cols - ms_error) / n))

  list(icc21 = icc21, icc2k = icc2k, n_targets = n, n_raters = k)
}

icc_input <- t(as.matrix(trial_matrix))
icc_stats <- icc_2way_random_absolute(icc_input)

category_pair_sd <- aggregate(
  cbind(sd, mean) ~ category_pair,
  data = observed_trial_summary,
  FUN = function(x) c(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE)
  )
)

category_pair_sd <- data.frame(
  category_pair = category_pair_sd$category_pair,
  sd_mean = category_pair_sd$sd[, "mean"],
  sd_median = category_pair_sd$sd[, "median"],
  sd_min = category_pair_sd$sd[, "min"],
  sd_max = category_pair_sd$sd[, "max"],
  mean_rating_mean = category_pair_sd$mean[, "mean"],
  mean_rating_median = category_pair_sd$mean[, "median"],
  row.names = NULL
)

catch_expectations <- c(slider_catch1 = "left", slider_catch2 = "right", slider_catch3 = "left", slider_catch4 = "right")
catch_scores <- catch_matrix
catch_scores$session_id <- results_raw$session_id
catch_scores$iteration <- results_raw$iteration
catch_scores$prolific_code <- results_raw$prolific_code

score_catches <- function(threshold) {
  correct <- cbind(
    catch_matrix$slider_catch1 <= threshold,
    catch_matrix$slider_catch2 >= (101 - threshold),
    catch_matrix$slider_catch3 <= threshold,
    catch_matrix$slider_catch4 >= (101 - threshold)
  )
  rowSums(correct, na.rm = TRUE)
}

thresholds <- c(1, 3, 5, 10, 15, 20)
catch_threshold_table <- do.call(
  rbind,
  lapply(thresholds, function(threshold) {
    scores <- score_catches(threshold)
    data.frame(
      threshold = threshold,
      pass_rule = paste0("4/4 within ", threshold, " points of the correct end"),
      n_pass = sum(scores == 4, na.rm = TRUE),
      n_fail = sum(scores < 4, na.rm = TRUE)
    )
  })
)

catch_participant_table <- data.frame(
  session_id = results_raw$session_id,
  iteration = results_raw$iteration,
  prolific_code = results_raw$prolific_code,
  slider_catch1 = catch_matrix$slider_catch1,
  slider_catch2 = catch_matrix$slider_catch2,
  slider_catch3 = catch_matrix$slider_catch3,
  slider_catch4 = catch_matrix$slider_catch4,
  correct_within_10 = score_catches(10),
  correct_within_5 = score_catches(5)
)

overall_summary <- data.frame(
  metric = c(
    "Participants in export",
    "Participants with complete catch data",
    "Observed rated trials in this export",
    "Unique blend images in this export",
    "Ratings per observed trial",
    "Overall mean rating",
    "ICC(2,1) across observed trials",
    "ICC(2,k) across observed trials"
  ),
  value = c(
    nrow(results_raw),
    sum(stats::complete.cases(catch_matrix)),
    length(trial_cols),
    nrow(observed_trial_summary),
    if (length(unique(observed_trial_summary$n)) == 1) unique(observed_trial_summary$n) else paste(range(observed_trial_summary$n), collapse = "-"),
    round(mean(long_trials$rating), 2),
    round(icc_stats$icc21, 3),
    round(icc_stats$icc2k, 3)
  )
)
```

## Overview

This report reads [results.json](`r results_path`) and summarizes the `blendScenes_prolific_A` export. This export contains ratings for 50 trials: `1, 5, 9, ..., 197`. Metadata are joined from [Blends.xlsx](`r blends_path`) using `id_blendu`, which matches the collage numbering used in the local `images/` folder.

```{r}
kable(overall_summary, col.names = c("Metric", "Value"))
```

The ICC shown here is an overall reliability estimate across the observed trials, treating participants as raters and trials as targets. A separate ICC for each individual trial is not identifiable from this design, because ICC needs variation across multiple targets rather than a single item. For each trial subsection below, the practical item-level summaries are the mean, SD, median, and histogram.

## SD by Category Pair

Because trial-specific ICC is not estimable from a single image, the summary figure below uses the per-image rating SD and aggregates it by category pair. Lower SD means participants were more tightly clustered for that type of blend.

```{r}
category_pair_sd_table <- category_pair_sd
category_pair_sd_table$sd_mean <- round(category_pair_sd_table$sd_mean, 2)
category_pair_sd_table$sd_median <- round(category_pair_sd_table$sd_median, 2)
category_pair_sd_table$sd_min <- round(category_pair_sd_table$sd_min, 2)
category_pair_sd_table$sd_max <- round(category_pair_sd_table$sd_max, 2)
category_pair_sd_table$mean_rating_mean <- round(category_pair_sd_table$mean_rating_mean, 2)
category_pair_sd_table$mean_rating_median <- round(category_pair_sd_table$mean_rating_median, 2)
kable(category_pair_sd_table, col.names = c("Category Pair", "Mean SD", "Median SD", "Min SD", "Max SD", "Mean Rating", "Median Rating"))
```

```{r, fig.width=8, fig.height=4.5}
ggplot(category_pair_sd, aes(x = reorder(category_pair, sd_mean), y = sd_mean, fill = category_pair)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", sd_mean)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  expand_limits(y = max(category_pair_sd$sd_mean, na.rm = TRUE) + 2) +
  labs(
    x = "Category pair",
    y = "Mean within-image SD",
    title = "Rating dispersion by category pair"
  ) +
  theme_minimal(base_size = 11)
```

## Catch Trials

The four catch trials appear to have the intended answers at the scale endpoints:

- `slider_catch1`: left end
- `slider_catch2`: right end
- `slider_catch3`: left end
- `slider_catch4`: right end

The table below shows how many participants would pass if you require all four catch trials to be within a given distance from the correct endpoint.

```{r}
kable(catch_threshold_table, col.names = c("Threshold", "Rule", "Pass", "Fail"))
```

Suggested cutoffs:

- Strict default: require all `4/4` catch trials within `10` points of the correct endpoint, and exclude participants with missing catch responses.
- Moderate alternative: require all `4/4` within `15` or `20` points if you want to allow slight slider imprecision near the endpoints.
- I would avoid a lenient `3/4` rule here unless you have a strong reason, because the study instructions explicitly told participants to choose the correct end of the scale.

```{r}
kable(catch_participant_table)
```

## Trial Summary

```{r}
trial_summary_table <- observed_trial_summary[, c("trial_id", "blend_name", "category_pair", "n", "mean", "sd", "median", "min", "max")]
trial_summary_table$mean <- round(trial_summary_table$mean, 2)
trial_summary_table$sd <- round(trial_summary_table$sd, 2)
trial_summary_table$median <- round(trial_summary_table$median, 2)
kable(trial_summary_table)
```

## Image Browser

```{r, results='asis', fig.width=7, fig.height=4}
for (trial in observed_trial_summary$trial_id) {
  trial_info <- observed_trial_summary[observed_trial_summary$trial_id == trial, ][1, ]
  image_path <- file.path(images_dir, paste0(trial, "_kolaz.png"))

  cat("\n\n### Trial ", trial, " (", trial_info$blend_name, ")\n\n", sep = "")
  cat("<p><img src=\"", image_path, "\" alt=\"Trial image\" style=\"width:70%;\" /></p>\n\n", sep = "")

  trial_ratings <- long_trials$rating[long_trials$trial_id == trial]
  cat(
    "**Category pair** = ", trial_info$category_pair,
    "; **category1** = ", trial_info$category1,
    "; **category2** = ", trial_info$category2,
    "\n\n",
    sep = ""
  )
  cat(
    "**Source images** = `", trial_info$img_name1, "` and `", trial_info$img_name2, "`\n\n",
    sep = ""
  )

  stats_line <- paste0(
    "**n** = ", length(trial_ratings),
    "; **mean** = ", round(mean(trial_ratings), 2),
    "; **sd** = ", round(stats::sd(trial_ratings), 2),
    "; **median** = ", round(stats::median(trial_ratings), 2),
    "; **range** = ", min(trial_ratings), "-", max(trial_ratings)
  )
  cat(stats_line, "\n\n")

  plot_df <- data.frame(rating = trial_ratings)
  p <- ggplot(plot_df, aes(x = rating)) +
    geom_histogram(binwidth = 5, boundary = 0, fill = "#3b6fb6", color = "white") +
    geom_vline(xintercept = mean(trial_ratings), color = "#c0392b", linewidth = 0.9) +
    scale_x_continuous(limits = c(1, 100), breaks = seq(0, 100, by = 10)) +
    labs(
      x = "Rating (1 = left end, 100 = right end)",
      y = "Count",
      title = paste("Histogram of ratings for trial", trial)
    ) +
    theme_minimal(base_size = 11)
  print(p)
}
```
