---
title: "blendScenes Typicality and Similarity Report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(readr)
library(readxl)
library(tidyr)
library(tibble)

workbook_path <- "../data/formr_downloads/blendScenes_prolific_A/results_version_A_long_with_images_and_paths.xlsx"
feature_dir <- "../data/deep_features/version_A_places365_alexnet"
typicality_path <- "../data/deep_features/version_A_places365_alexnet/image_manifest_chatgpt_typicality.csv"
blends_path <- "../data/Blends.xlsx"

normalize_image_id <- function(x) {
  x <- as.character(x)
  x <- sub("\\.[A-Za-z0-9]+$", "", x)
  trimws(x)
}

cosine_pair <- function(a, b) {
  denom <- sqrt(sum(a * a)) * sqrt(sum(b * b))
  if (is.na(denom) || denom == 0) {
    return(NA_real_)
  }
  sum(a * b) / denom
}

l2_pair <- function(a, b) {
  sqrt(sum((a - b) ^ 2))
}

results_long <- read_excel(workbook_path, sheet = "long_results") |>
  filter(trial_type == "trial") |>
  mutate(
    blend_image_id = normalize_image_id(blend_image),
    left_image_id = normalize_image_id(left_image),
    right_image_id = normalize_image_id(right_image)
  )

trial_index <- results_long |>
  distinct(trial_id, blend_image, left_image, right_image, blend_image_id, left_image_id, right_image_id)

blend_lookup <- read_excel(blends_path) |>
  transmute(
    trial_id = as.integer(id_blendu),
    category1 = as.character(category1),
    category2 = as.character(category2),
    category_pair = paste0(category1, "-", category2)
  )

typicality_df <- read_csv(typicality_path, show_col_types = FALSE) |>
  mutate(image_id = normalize_image_id(image_name)) |>
  select(image_id, image_name, typ_chatgpt_m, typ_chatgpt_sd, typ_chatgpt_min, typ_chatgpt_max, typ_chatgpt_n, typ_chatgpt_model)

feature_files <- c("conv3_features.csv", "conv5_features.csv", "fc6_features.csv", "fc7_features.csv")

trial_side_list <- lapply(feature_files, function(file_name) {
  layer_name <- sub("_features\\.csv$", "", basename(file_name))
  feature_df <- read_csv(file.path(feature_dir, file_name), show_col_types = FALSE) |>
    mutate(image_id = normalize_image_id(image_name))

  feature_mat <- feature_df |>
    select(-image_name, -image_id) |>
    as.matrix()
  rownames(feature_mat) <- feature_df$image_id

  left_df <- trial_index |>
    transmute(
      trial_id,
      layer = layer_name,
      side = "left",
      blend_image,
      side_image = left_image,
      blend_image_id,
      side_image_id = left_image_id,
      cosine_similarity = map2_dbl(
        blend_image_id, left_image_id,
        ~ cosine_pair(feature_mat[.x, ], feature_mat[.y, ])
      ),
      l2_distance = map2_dbl(
        blend_image_id, left_image_id,
        ~ l2_pair(feature_mat[.x, ], feature_mat[.y, ])
      )
    )

  right_df <- trial_index |>
    transmute(
      trial_id,
      layer = layer_name,
      side = "right",
      blend_image,
      side_image = right_image,
      blend_image_id,
      side_image_id = right_image_id,
      cosine_similarity = map2_dbl(
        blend_image_id, right_image_id,
        ~ cosine_pair(feature_mat[.x, ], feature_mat[.y, ])
      ),
      l2_distance = map2_dbl(
        blend_image_id, right_image_id,
        ~ l2_pair(feature_mat[.x, ], feature_mat[.y, ])
      )
    )

  bind_rows(left_df, right_df)
})

trial_side <- bind_rows(trial_side_list) |>
  left_join(blend_lookup, by = "trial_id") |>
  left_join(typicality_df, by = c("side_image_id" = "image_id")) |>
  rename(
    side_typicality = typ_chatgpt_m,
    side_typicality_sd = typ_chatgpt_sd
  )

layer_summary <- trial_side |>
  group_by(layer) |>
  summarise(
    n_trial_sides = n(),
    cor_typicality_cosine = cor(side_typicality, cosine_similarity, use = "complete.obs"),
    cor_typicality_l2 = cor(side_typicality, l2_distance, use = "complete.obs"),
    cosine_lm_r2 = summary(lm(cosine_similarity ~ side_typicality))$r.squared,
    l2_lm_r2 = summary(lm(l2_distance ~ side_typicality))$r.squared,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

pair_summary <- trial_side |>
  group_by(layer, category_pair) |>
  summarise(
    n_trial_sides = n(),
    cor_typicality_cosine = cor(side_typicality, cosine_similarity, use = "complete.obs"),
    cor_typicality_l2 = cor(side_typicality, l2_distance, use = "complete.obs"),
    cosine_lm_r2 = summary(lm(cosine_similarity ~ side_typicality))$r.squared,
    l2_lm_r2 = summary(lm(l2_distance ~ side_typicality))$r.squared,
    mean_typicality = mean(side_typicality, na.rm = TRUE),
    mean_cosine = mean(cosine_similarity, na.rm = TRUE),
    mean_l2 = mean(l2_distance, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

trial_typicality <- trial_side |>
  select(trial_id, category_pair, side, side_typicality) |>
  group_by(trial_id, category_pair, side) |>
  summarise(side_typicality = mean(side_typicality, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(
    id_cols = c(trial_id, category_pair),
    names_from = side,
    values_from = side_typicality
  ) |>
  mutate(
    left = as.numeric(left),
    right = as.numeric(right),
    typicality_diff_right_minus_left = right - left,
    typicality_mean = (left + right) / 2
  )

blend_type_typicality_summary <- trial_typicality |>
  group_by(category_pair) |>
  summarise(
    n_trials = n(),
    mean_left_typicality = mean(left, na.rm = TRUE),
    mean_right_typicality = mean(right, na.rm = TRUE),
    mean_typicality_diff_right_minus_left = mean(typicality_diff_right_minus_left, na.rm = TRUE),
    sd_typicality_diff_right_minus_left = sd(typicality_diff_right_minus_left, na.rm = TRUE),
    mean_typicality = mean(typicality_mean, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

typicality_anova_tidy <- broom::tidy(aov(typicality_mean ~ category_pair, data = trial_typicality))
typicality_diff_anova_tidy <- broom::tidy(aov(typicality_diff_right_minus_left ~ category_pair, data = trial_typicality))

overall_summary <- tibble(
  metric = c(
    "Observed version A trials",
    "Trial-side rows",
    "Feature layers summarized",
    "Typicality source",
    "Similarity measures"
  ),
  value = c(
    n_distinct(trial_side$trial_id),
    nrow(trial_side),
    paste(sort(unique(trial_side$layer)), collapse = ", "),
    "ChatGPT image typicality means from image_manifest_chatgpt_typicality.csv",
    "Cosine similarity and Euclidean distance between the blend and each side image"
  )
)
```

## Overview

This report links ChatGPT-estimated image typicality from [image_manifest_chatgpt_typicality.csv](`r typicality_path`) to blend-to-side deep-feature similarity from the Places365 AlexNet features in [version_A_places365_alexnet](`r feature_dir`).

Each row in the analysis is one `trial x side x layer` combination. For each row, the report uses:

- the side image typicality estimate
- the cosine similarity between the blend and that side image
- the L2 distance between the blend and that side image
- the blend type (`category_pair`)

```{r}
kable(overall_summary, col.names = c("Metric", "Value"))
```

## Layer Summary

```{r}
kable(layer_summary, col.names = c("Layer", "Trial-Sides", "Cor: typicality vs cosine", "Cor: typicality vs L2", "Cosine model R2", "L2 model R2"))
```

## By Blend Type Summary

```{r}
kable(pair_summary, col.names = c("Layer", "Blend Type", "Trial-Sides", "Cor: typicality vs cosine", "Cor: typicality vs L2", "Cosine model R2", "L2 model R2", "Mean Typicality", "Mean Cosine", "Mean L2"))
```

## Typicality Differences by Blend Type

This section asks whether the side images paired with a blend differ in typicality depending on blend type, and whether the right-minus-left typicality imbalance differs across blend types.

```{r}
kable(blend_type_typicality_summary, col.names = c("Blend Type", "Trials", "Mean Left Typicality", "Mean Right Typicality", "Mean Right-Left Typicality Diff", "SD Right-Left Typicality Diff", "Mean Typicality"))
```

```{r}
kable(typicality_anova_tidy)
```

```{r}
kable(typicality_diff_anova_tidy)
```

## Typicality Level by Blend Type

```{r, fig.width=10, fig.height=4.8}
ggplot(trial_typicality, aes(x = category_pair, y = typicality_mean, fill = category_pair)) +
  geom_boxplot(alpha = 0.8, width = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.12, alpha = 0.7, size = 1.8, show.legend = FALSE) +
  labs(
    x = "Blend type",
    y = "Mean side-image typicality",
    title = "Typicality level by blend type"
  ) +
  theme_minimal(base_size = 11)
```

## Left-Right Typicality Imbalance by Blend Type

```{r, fig.width=10, fig.height=4.8}
ggplot(trial_typicality, aes(x = category_pair, y = typicality_diff_right_minus_left, fill = category_pair)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey70") +
  geom_boxplot(alpha = 0.8, width = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.12, alpha = 0.7, size = 1.8, show.legend = FALSE) +
  labs(
    x = "Blend type",
    y = "Right minus left typicality",
    title = "Left-right typicality imbalance by blend type"
  ) +
  theme_minimal(base_size = 11)
```

## Cosine Similarity vs Typicality

```{r, fig.width=8, fig.height=4.8}
ggplot(trial_side, aes(x = side_typicality, y = cosine_similarity, color = layer)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    x = "ChatGPT-estimated side-image typicality",
    y = "Blend-to-side cosine similarity",
    title = "Typicality and cosine similarity across layers"
  ) +
  theme_minimal(base_size = 11)
```

## L2 Distance vs Typicality

```{r, fig.width=8, fig.height=4.8}
ggplot(trial_side, aes(x = side_typicality, y = l2_distance, color = layer)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    x = "ChatGPT-estimated side-image typicality",
    y = "Blend-to-side L2 distance",
    title = "Typicality and L2 distance across layers"
  ) +
  theme_minimal(base_size = 11)
```

## Cosine by Blend Type

```{r, fig.width=12, fig.height=7}
ggplot(trial_side, aes(x = side_typicality, y = cosine_similarity, color = side)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(category_pair ~ layer, scales = "free") +
  labs(
    x = "ChatGPT-estimated side-image typicality",
    y = "Blend-to-side cosine similarity",
    title = "Typicality and cosine similarity split by blend type and layer"
  ) +
  theme_minimal(base_size = 10)
```

## L2 by Blend Type

```{r, fig.width=12, fig.height=7}
ggplot(trial_side, aes(x = side_typicality, y = l2_distance, color = side)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(category_pair ~ layer, scales = "free") +
  labs(
    x = "ChatGPT-estimated side-image typicality",
    y = "Blend-to-side L2 distance",
    title = "Typicality and L2 distance split by blend type and layer"
  ) +
  theme_minimal(base_size = 10)
```

## Trial-Side Table

```{r}
trial_table <- trial_side |>
  select(layer, category_pair, trial_id, side, blend_image, side_image, side_typicality, cosine_similarity, l2_distance, side_typicality_sd) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(trial_table)
```
